Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2022-01-18 analysis/paper/London_HauloutBeahvior.Rmd
JUSTIN C FINAL COMMENTS ADDRESSED

  - name: Justin Crawford

@@

  occur in spring (April-June) as sea ice begins to melt and retreat northward.

@@

Haul-out behavior data from bio-loggers deployed on 
`r dplyr::n_distinct(deploy_table$speno)` bearded, ribbon, and spotted seals were
subset to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
following annual molt, which, depending on deployment date, limits the duration
of haul-out data they provide. Additionally, bio-loggers attached to the head or
dorsal region are often dry while the seal is at the surface and can slightly
bias the hourly percent-dry values reported by the bio-logger. For this study,
in cases where both bio-logger types were deployed, hourly percent-dry
observations from the flipper tag were preferred.

@@

```{r tblDeploy, include=TRUE, tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020. Total seal hours represents the sum of available data across all seals. Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table"}

@@

Tags that fall off due to molt, attachment failure, or seal mortality and remain
on ice or land can still send data to satellites; i.e., the bio-logger will be
on ice or land and dry, therefore, it will record and transmit data suggesting
the seal is hauled out. As such, end times of each deployment were identified by
examining bio-logger locations and dive behavior to determine when bio-loggers
ceased providing data consistent with seal behavior. Data outside of the
deployment start and end times were discarded prior to analysis. After
approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
reported implausible hourly percent-dry data. All data after the first instance
of unrealistic values were censored from this analysis. Figure
\@ref(fig:dataCal) shows the distribution of all haul-out data
across the study season for each species. Observations for ribbon and spotted
seals are concentrated in the months of May and June due to the timing of
deployment (April and May) and the timing of molt (May and June). During molt,
seals (and their attached bio-loggers) spend more time out of the water and more
data are transmitted. Molt timing also impacts when many deployments end as any
bio-loggers adhered to the hair will fall off. Bearded seal deployments are more
evenly spread with fewer observations in May and June. The majority of bearded
seal deployments start later in the summer and by May bio-loggers have either
fallen off or the batteries are depleted.

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and animal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m [@lopez2013]; we chose 2500m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50m. Any days where haul-out observations were
present but location data were missing we used the last calculated weighted
average daily location, and any days where the location intersected with land
were removed from the data set. We recognized that bearded and spotted seals
haul out on land. However, assessing the relationship between haul-out behavior
and weather covariates and their availability for aerial surveys on land was
outside the scope of this study. Additionally, any daily locations on land are
likely more reflective of coordinate averaging and measurement error instead of
actual use of coastal features. Figure \@ref(fig:dataMap) shows the spatial
distribution weighted locations with available haul-out behavior data for each
species across the study area.

@@

```{r dataMap, fig.height = 3.75, fig.id="dataMap", include=TRUE, fig.cap="Spatial distribution of haul-out data during the months of March through July 15 for each of the three species. Data were collated across all years between 2005 and 2020."}

@@

ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(
    data = locs_count, size = 0.125,
    aes(fill = ct)
  ) +
  scale_fill_gradientn(colors = rev(met.brewer("Greek")),
                    trans = "log10", aesthetics = "fill",
    name = "number of locations",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
                           barheight = 0.5, title.hjust = 0.5)) + 
  facet_grid(~ species) +
  theme(legend.position = "bottom") +
  ggtitle("Spatial distribution of haul-out behavior data")
```

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated
using the {solaR} package [@perpinan2012] within the R statistical environment
[@rcoreteam2021] using the daily locations. We included several weather
variables shown to affect haul-out behavior in other Arctic pinnipeds
[@reder2003; @udevitz2009; @perry2017]. In particular, we linked the weighted
average daily locations from our bio-loggers to key weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America.
Numerous weather variables are made available across the region 8 times daily.
For this study, NARR weather values were subset to the extent of our study area
over the Bering and Chukchi seas at 3-hr intervals based on the native grid
resolution of 32 km (1024 sq. km). The following meteorological variables were
interpolated and assigned to seal locations using a bilinear method: 1) air
temperature at 2m above the earth's surface, 2) wind consisting of northerly and
easterly vector components converted to wind speed using the Euclidean norm, 3)
barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

@@

covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "young-of-the-year, subadult, adult male and adult female",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "local solar hour using 6 variables of a Fourier-series basis",
  "Day",   "Continuous", "Bio-logger", 
  "linear, quadratic, and cubic effects of day-of-year",
  "Precip", "Continuous", "NARR", 
  "convective precipitation (kg/m^2^)",
  "Pressure", "Continuous", "NARR", 
  "atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "air temperatures at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "northerly and easterly vector components for wind converted into a single wind speed via the Euclidean norm",
  "Northing", "Continuous", "Bio-logger","latitude divided by the mean latitude across all locations (for bearded seals only)",
  "Year", "Continuous", "Bio-logger", "For the set of models examining inter-annual variation in sea ice use, we fitted models with the addition of year by day-of-year interactions." 
)

@@

A typical model fitting exercise would also include a model selection process.
However, AIC (and similar criteria) is not suitable when using
pseudo-likelihoods, because pseudo-data generated in the model fitting process
[@verhoef2010a] differ between models [@teneyck2018]. After fitting GLMPM
models, we instead used "type III" $F$-tests to calculate $p$-values
[@verhoef2010a] to evaluate model performance and important terms. We also
produced predictions of haul-out behavior as a function of three influential
predictors (e.g. solar hour, day-of-year, age-sex). Weather covariates for these
predictions were based on daily or hourly smoothed weather covariate values
across the study region. Such predictions were then used to develop haul-out
probability surfaces (see figures \@ref(beardedHOCal), \@ref(ribbonHOCal), and
\@ref(spottedHOCal)), explore conditional effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out
behavior was further used to regress against the annual maximum sea ice extent
in the study region.

@@

```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.cap="Predicted haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}

@@

Ribbon seals show a pattern of gradually increasing haul-out probability
in April that peaks in late May for subadults and in early June for
adults (Figure \@ref(fig:ribbonHOCal)). The behavior is clearly centered
around local solar noon and expands to other hours later in the season
as seals enter their molting period. Subadults showed an earlier start
and more intense haul-out activity in April and May. The
young-of-the-year records begin after weaning and the model predictions
demonstrate the ontogeny of in-water activities (e.g. diving, foraging)
in May. Adult females have a more protracted haul-out season compared to
males, and more time spent resting in June and July. The model suggests
adult male ribbon seals complete their molt by the end of June. Lastly,
for adults of both sexes there is some indication for a shift to a
crepuscular haul-out pattern in late June and July.

```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}

@@

```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}

@@

Unlike previous analyses of seal haul-out behavior (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class on haul-out
probabilities for all species except bearded seals because of low sample size.
Although ribbon and spotted seals exhibited a unimodal diel haul-out pattern
generally centered around local solar noon, there were key differences across
species, age, and sex that match expectations given what we know of their
ecological behavior. Spotted seals are known to form triads during the breeding
season where a female and dependent pup are accompanied on the ice by a suitor
male [@frost2018]. The male waits for the female to wean the pup and enter
estrus, and fends off any other potential suitor males. Triad formation results
in both males and females spending a large portion of the day hauled out on ice
and a protracted spring haul-out season for both sexes. While females are still
nursing the pup and not yet in estrus they may be less inclined to interrupt
their haul-out and enter the water where the suitor male would attempt mating.
We see this reflected in the predicted haul-out behavior, with both males and
females exhibiting a broad distribution of time out of the water throughout the
solar day and the season. Ribbon seals are not known to form triads and our
model predicts a progression of increased haul-out behavior with females
starting earlier in the season than males. Notably, female ribbon seals spend a
large portion of the day in the water during the pupping period, aligning with
the hypothesis that ribbon seal females continue foraging while nursing.
Subadult ribbon and spotted seals begin consistent haul-out behavior earlier and
follow a pattern seen in other phocids where yearlings and subadults molt first
followed by adult females and males [@reder2003, @thompson1987, @kirkman2003].

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g. Udevitz et al.
[-@udevitz2009]) and a few select studies of ice-associated seals
[@hamilton2018, @perry2017]. Ribbon seals seemed to be the most influenced by
weather, with wind, temperature, barometric pressure, and precipitation all
being important components of the model. Spotted seals were most affected by
wind and barometric pressure. And, for bearded seals, the model indicated wind
and temperature had the greatest impact. In general, and as might be expected,
seals were more likely to haul-out when daily temperatures were warmer, winds
speeds were lower, barometric pressure was higher, and precipitation was lower.
These weather conditions are general indicators of increased solar radiation
which provides energetic benefits. Low winds and precipitation may also
provide better predator detection conditions. Our results highlight the
importance of incorporating weather covariates when analyzing haul-out 
behavior and calculating availability corrections for aerial surveys.

Our models detected small deviations in the timing and magnitude of annual peaks
in haul-out behavior for ribbon and spotted seals. The timing of peak haul-out
activity appears to fall within a relatively narrow time window of 3-4 weeks in
late May and early June. This consistency across 15 years is likely a reflection
of the relationship between a critical photoperiod and the timing of important
life history stages [@bronson2009; @temte1994]. However, along with a critical
photoperiod, ribbon and spotted seals are dependent upon the presence of sea ice
for pupping and molt. We didn't find any support in our models for a
relationship between the timing of peaks in haul-out behavior and the amount of
yearly maximum sea ice. This could indicate that, while the extent of spring sea
ice in the Bering sea varied widely during our study period, seals were still
able to locate sea ice and haul out. We should expect, however, that some
minimal threshold in the spatial extent or density of sea ice will have a
meaningful impact on the timing of peak haul-out behavior. Additionally, while
from an ecological perspective the haul-out behavior appears consistent, the
interannual differences in timing and magnitude are large enough to have
important ramifications on calculations of abundance and trend. Those
ramifications will only be exacerbated if climate variability amplifies
interannual differences.

@@

We recognize that the species and ecosystems studied in this paper are within
the traditional range of the Inpuiaq and Yupik people. And, we would like to
acknowledge and thank them as the original stewards of these ecosystems. The
deployment of bio-logging devices used in this study were often done in
collaboration with Alaska Native seal hunters near their communities. We would
like to especially acknowledge the communities of Kotzebue, Koyuk, Nuiqsut,
Utqiaġvik, and Ulguniq (Wainwright) and the following individuals: James Adams,
Jeff Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris,
Grover Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry
Goodwin, John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor,
Virgil Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen Stone, and Randy
Toshavik from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Billy Adams,
Mary-Ellen Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman,
Howard Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik,
Shawn Oktollik, Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer,
Bob Shears, and Joe Skin from the North Slope Borough, Alaska;

We are grateful for the assistance in catching and sampling seals by
James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir
Burkanov, Cynthia Christman, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey
Goldstein, Jeff Harris, Markus Horning, John Jansen, Shawn Johnson,
Charles Littnan, Lloyd Lowry, Erin Moreland, Mark Nelson, Lorrie Rea, Gay Sheffield,
Brent Stewart, and Dave Withrow. We also appreciate the commitment to
science and safety by all officers and crew of the NOAA Ship Oscar
Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.

>>>>

  - name: Justin A. Crawford

@@

  occur in spring (April -- June) as sea ice begins to melt and retreat northward.

@@

```{r deploystats}
species_n <- deploy_table %>% 
  group_by(species) %>% 
  distinct(speno) %>% 
  count(species)
```

Haul-out behavior data from 
`r species_n %>% .$n %>% sum()` bio-loggers deployed on 
`r species_n %>% filter(species == "Bearded seal") %>% .$n` bearded, 
`r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals were
subset to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
following annual molt, which, depending on deployment date, limits the duration
of haul-out data they provide. Additionally, bio-loggers attached to the head or
dorsal region are often dry while the seal is at the surface and can slightly
bias the hourly percent-dry values reported by the bio-logger. For this study,
in cases where both bio-logger types were deployed, hourly percent-dry
observations from the flipper tag were preferred.

@@

```{r tblDeploy, include=TRUE, tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020. Total seal hours represents the sum of hourly observations across all seals used in the analysis. Because young-of-the-year are advanced to subadult on 1 March of the following year, some individual seals are represented in both columns in this table"}

@@

Tags that fall off due to molt, attachment failure, or seal mortality and remain
on ice or land can still send data to satellites; i.e., the bio-logger will be
on ice or land and dry, therefore, it will record and transmit data suggesting
the seal is hauled out. As such, end times of each deployment were identified by
examining bio-logger locations, percent-dry records, and dive behavior (if
available) to determine when bio-loggers ceased providing data consistent with
seal behavior. Data outside of the deployment start and end times were discarded
prior to analysis. After approximately 9 months, 7 devices deployed on the rear
flipper of bearded seals reported implausible hourly percent-dry data. All data
after the first instance of unrealistic values were censored from this analysis.
Figure \@ref(fig:dataCal) shows the distribution of all haul-out data across the
study season for each species. Observations for ribbon and spotted seals are
concentrated in the months of May and June due to the timing of deployment
(April and May) and the timing of molt (May and June). During molt, seals (and
their attached bio-loggers) spend more time out of the water and more data are
transmitted. Molt timing also impacts when many deployments end as any
bio-loggers adhered to the hair will fall off. Bearded seal deployments are more
evenly spread with fewer observations in May and June. The majority of bearded
seal deployments start later in the summer and by May bio-loggers have either
fallen off or their batteries are depleted.

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and animal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m [@lopez2013]; we chose 2500m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50m. Any days where haul-out observations were
present but location data were missing we used the last calculated weighted
average daily location, and any days where the location intersected with land
were removed from the data set. We recognized that bearded and spotted seals
haul out on land. However, assessing the relationship between haul-out behavior
and weather covariates and their availability for aerial surveys on land was
outside the scope of this study. Additionally, any daily locations on land are
likely more reflective of coordinate averaging and measurement error instead of
actual use of coastal features. Figure \@ref(fig:dataMap) shows the spatial
distribution of weighted locations with available haul-out behavior data for each
species across the study area.

@@

(ref:mapCaption) Spatial distribution of haul-out data during the months of March through July 15 for each of the three species. Data were collated across all years between 2005 and 2020. Each hexagonal cell represents an area of 50 km^2^

```{r dataMap, fig.height = 3.75, include=TRUE, fig.cap="(ref:mapCaption)"}

@@

ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(
    data = locs_count, size = 0.125,
    aes(fill = ct)
  ) +
  scale_fill_gradientn(colors = rev(met.brewer("Greek")),
                    trans = "log10", aesthetics = "fill",
    name = "number of locations",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
                           barheight = 0.5, title.hjust = 0.5)) + 
  facet_grid(~ species) +
  theme(legend.position = "bottom") +
  ggtitle("Spatial distribution of haul-out behavior data") +
  labs(caption = bquote("One hexagonal cell represents 50"~km^2))
```

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated
using the {solaR} package [@perpinan2012] within the R statistical environment
[@rcoreteam2021] using the daily locations. We also linked the weighted average
daily locations from our bio-loggers to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America.
Numerous weather variables are made available across the region 8 times daily.
For this study, NARR weather values were subset to the extent of our study area
over the Bering and Chukchi seas at 3-hr intervals based on the native grid
resolution of 32 km (1024 sq. km). The following meteorological variables are
known to affect haul-out behavior in other Arctic pinnipeds [@reder2003;
@udevitz2009; @perry2017] and were interpolated and assigned to daily seal locations
using a bilinear method: 1) air temperature at 2m above the earth's surface, 2)
wind consisting of northerly and easterly vector components converted to wind
speed using the Euclidean norm, 3) barometric pressure at sea level, and 4)
precipitation (Table \@ref(tab:tblCovariates)).

@@

(ref:kgm2) kg/m^2^

@@

covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "young-of-the-year, subadult, adult male and adult female",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "local solar hour using 6 variables of a Fourier-series basis",
  "Day",   "Continuous", "Bio-logger", 
  "linear, quadratic, and cubic effects of day-of-year",
  "Precip", "Continuous", "NARR", 
  "convective precipitation (ref:kgm2)",
  "Pressure", "Continuous", "NARR", 
  "atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "air temperatures at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "northerly and easterly vector components for wind converted into a single wind speed via the Euclidean norm",
  "Northing", "Continuous", "Bio-logger","latitude divided by the mean latitude across all locations (for bearded seals only)",
  "Year", "Continuous", "Bio-logger", "For the set of models examining inter-annual variation in sea ice use, we fitted models with the addition of year by day-of-year interactions." 
)

@@

A typical model fitting exercise would also include a model selection process.
However, AIC (and similar criteria) is not suitable when using
pseudo-likelihoods, because pseudo-data generated in the model fitting process
[@verhoef2010a] differ between models [@teneyck2018]. After fitting GLMPM
models, we instead used "type III" $F$-tests to calculate $p$-values
[@verhoef2010a] to evaluate model performance and important terms. We also
produced predictions of haul-out behavior as a function of three influential
predictors (e.g. solar hour, day-of-year, age-sex). Weather covariates for these
predictions were based on daily or hourly smoothed weather covariate values
across the study region. Such predictions were then used to develop haul-out
probability surfaces (see figures \@ref(fig:beardedHOCal), \@ref(fig:ribbonHOCal), and
\@ref(fig:spottedHOCal)), explore conditional effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out
behavior was further used to regress against the annual maximum sea ice extent
in the study region.

@@

```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes. Hour of the day was fixed at local solar noon."}

@@

Ribbon seals exhibited a pattern of gradually increasing haul-out probability
in April that peaks in late May for subadults and in early June for
adults (Figure \@ref(fig:ribbonHOCal)). The behavior is clearly centered
around local solar noon and expands to other hours later in the season
as seals enter their molting period. Subadults showed an earlier start
and more intense haul-out activity in April and May. The
young-of-the-year records begin after weaning and the model predictions
demonstrate the ontogeny of in-water activities (e.g. diving, foraging)
in May. Adult females have a more protracted haul-out season compared to
males, and more time spent resting in June and July. The model suggests
adult male ribbon seals complete their molt by the end of June. Lastly,
for adults of both sexes there is some indication for a shift to a
crepuscular haul-out pattern in late June and July.

```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes. Hour of the day was fixed at local solar noon."}

@@

```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes. Hour of the day was fixed at local solar noon."}

@@

Unlike previous analyses of seal haul-out behavior (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class on haul-out
probabilities for all species except bearded seals because of low sample size.
Although ribbon and spotted seals exhibited a unimodal diel haul-out pattern
generally centered around local solar noon, there were key differences across
species, age, and sex that match our understandinf of their
ecological behavior. Spotted seals are known to form triads during the breeding
season where a female and dependent pup are accompanied on the ice by a suitor
male [@frost2018]. The male waits for the female to wean the pup and enter
estrus, and fends off any other potential suitor males. Triad formation results
in both males and females spending a large portion of the day hauled out on ice
and a protracted spring haul-out season for both sexes. While females are still
nursing the pup and not yet in estrus they may be less inclined to interrupt
their haul out and enter the water where the suitor male would attempt mating.
We see this reflected in the predicted haul-out behavior, with both males and
females exhibiting a broad distribution of time out of the water throughout the
solar day and the season. Ribbon seals are not known to form triads and our
model predicts a progression of increased haul-out behavior with females
starting earlier in the season than males. Notably, female ribbon seals spend a
large portion of the day in the water during the pupping period, aligning with
the hypothesis that ribbon seal females continue foraging while nursing.
Subadult ribbon and spotted seals begin consistent haul-out behavior earlier and
follow a pattern seen in other phocids where yearlings and subadults molt first
followed by adult females and males [@reder2003, @thompson1987, @kirkman2003].

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g. Udevitz et al.
[-@udevitz2009]) and a few select studies of ice-associated seals
[@hamilton2018, @perry2017]. Ribbon seals seemed to be the most influenced by
weather, with wind, temperature, and barometric pressure all
being important components of the model. Spotted seals were most affected by
wind and barometric pressure. And, for bearded seals, the model indicated wind
and temperature had the greatest impact. In general, and as might be expected,
seals were more likely to haul-out when daily temperatures were warmer, winds
speeds were lower, barometric pressure was higher, and precipitation was lower.
These weather conditions are general indicators of increased solar radiation
which provides energetic benefits. Low winds and precipitation may also
provide better predator detection conditions. Our results highlight the
importance of incorporating weather covariates when analyzing haul-out 
behavior and calculating availability corrections for aerial surveys.

Our models detected small deviations in the timing and magnitude of annual peaks
in haul-out behavior for ribbon and spotted seals. The timing of peak haul-out
activity appears to fall within a relatively narrow time window of 3-4 weeks in
late May and early June. This consistency across 15 years is likely a reflection
of the relationship between a critical photoperiod and the timing of important
life history stages [@bronson2009; @temte1994]. However, along with a critical
photoperiod, ribbon and spotted seals are dependent upon the presence of sea ice
for pupping and molt. We didn't find any support in our models for a
relationship between the timing of peaks in haul-out behavior and the amount of
yearly maximum sea ice. This could indicate that, while the extent of spring sea
ice in the Bering sea varied widely during our study period, seals were still
able to locate sea ice and haul out. We should expect, however, that some
minimal threshold in the spatial extent or density of sea ice will have a
meaningful impact on the timing of peak haul-out behavior --- if there is no sea
ice, seals will not haul out or be forced to use terrestrial haul-out sites.
Additionally, while from an ecological perspective the haul-out behavior appears
consistent, the interannual differences in timing and magnitude are large enough
to have important ramifications on calculations of abundance and trend. Those
ramifications will only be exacerbated if climate variability amplifies
interannual differences.

@@

We recognize that the species and ecosystems studied in this paper are within
the traditional range of the Inpuiaq and Yupik people. And, we would like to
acknowledge and thank them as the original stewards of these ecosystems. The
deployment of bio-logging devices used in this study were often done in
collaboration with Alaska Native seal hunters near their communities. We would
like to especially acknowledge the communities of Kotzebue, Koyuk, Nome,
Nuiqsut, Scammon Bay, St. Michael, Utqiaġvik, and Ulguniq (Wainwright) and the
following individuals: James Adams, Jeff Barger, David Barr, Wendell Booth,
Cyrus Harris, Nereus 'Doc' Harris, Grover Harris, Lee Harris, Tom Jones, Frank
Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin, Pearl Goodwin, Willie
Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan
Savetilik, Ross Schaeffer, Allen Stone, and Randy Toshavik from Kotzebue,
Alaska; Merlin Henry from Koyuk, Alaska; Tom Gray from Nome; Vernon Long and
Richard Tukle from Nuiqsuit, Alaska; Morgan Simon, River Simon, and Al Smith
from Scammon Bay, Alaska; Alex Niksik Jr. from St. Michael, Alaska; Billy Adams,
Mary-Ellen Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman,
Howard Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik,
Shawn Oktollik, Stacey Osborn, Fred Rexford, Chuck Schaeffer, Bob Shears, and
Joe Skin from the North Slope Borough, Alaska.

We are grateful for the assistance in catching and sampling seals by Ryan Adam,
James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov,
Cynthia Christman, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb Fauquier, Shannon
Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff Harris, Markus
Horning, John Jansen, Shawn Johnson, Charles Littnan, Lloyd Lowry, Erin
Moreland, Mark Nelson, Justin Olnes, Lorrie Rea, Gay Sheffield, Brent Stewart,
and Dave Withrow. We also appreciate the commitment to science and safety by all
officers and crew of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and
the RV Thomas G. Thompson.
